Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-10-14 Manuscript.tex
Published

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Palaeo3}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

@@

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={Department of Evolutionary Biology,
Ecology and Environmental Science, Biodiversity Research Institute
(IRBio), University of Barcelona},addressline={Avinguda Diagonal,
643},city={Barcelona},country={Spain},countrysep={,},postcode={8028},postcodesep={}}

@@

\begin{abstract}
This study provides a reassessment of Magnesium to Calcium (Mg/Ca)
ratios in Atlantic limpet shells to determine past sea surface
temperatures (SST). While \emph{Patella depressa} along the Spanish
shoreline has repeatedly produced reliable correlations between SST and
Mg/Ca ratios, this relationship is not the case for other patelloid
species. \emph{Patella vulgata}, \emph{Nacella deaurata}, and \emph{N.
magellanica} have been studied using Mg/Ca ratios with mixed or contrary
results. In this study, we present elemental maps of these three species
together with oxygen isotope ratios (\(\delta\)\textsuperscript{18}O)
that confirm a good relationship with SST. Our dataset also reassesses a
specimen which was previously unsuccessful in providing significant
correlations between \(\delta\)\textsuperscript{18}O values and Mg/Ca
ratios. By reassessing these species and including modern and
archaeological specimens (n=12) from three patelloid species (\emph{P.
vulgata}, \emph{N. deaurata}, and \emph{N. magellanica}) we further add
to the growing set of evidence for the reliable use of Mg/Ca ratios to
detect palaeotemperature change. As a result, these species can in the
future serve to determine ontogenetic age and season of capture as well
as to reveal locations of interest within the growth record (i.e.~annual
temperature minima and maxima) for the targeted analysis using
\(\delta\)\textsuperscript{18}O and clumped isotope analysis.
\end{abstract}

@@

Limpet shells are commonly found within archaeological sites and past
shorelines due to their robust carbonate structure and the long-term use
of limpets as a marine food source
\citep[\citet{Robson2023-oz}]{Alvarez2011-rm}. They have been
successfully studied in the context of coastal subsistence economies and
site occupation on seasonal
\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
scales \citep{Ortiz2015-mr}, as well as archives of palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change recorded in
patelloid limpet shells largely relies on the measurement of
\(\delta\)\textsuperscript{18}O values within the calcitic parts of the
shell, but attempts have been made to also use elemental ratios, such as
magnesium to calcium (Mg/Ca) ratios
\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative measure
that potentially provides a more accurate SST estimate than
\(\delta\)\textsuperscript{18}O-values,as the latter proxy is affected
by changes in the mixing of fresh- and saltwater. In addition, the data
acquisition of elemental ratios --- either through
laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
induced breakdown spectroscopy (LIBS), can be much faster and
cost-effective, increasing the number of specimens that can be studied
overall \citep{Durham2017-fh, Hausmann2023-ih}.

@@

Recent research particularly of \emph{Patella} spp. in the Mediterranean
and southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
However, there remains a lack of clarity for Atlantic limpet species,
particularly since it was shown, that the Mg/Ca ratio in their shells
are not reliable recorders of palaeotemperature \citep{Graniero2017-io}.

In this study we will repeat and expand the analysis of Atlantic limpets
to determine the reliability of Mg/Ca ratios as palaeotemperature
proxies. To do this, we sampled a set of previously published and
unpublished limpet specimens dating to modern and archaeological
contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
shell sections, which allows navigation of the complex elemental
structure of the shell and better separate the external,
temperature-related changes from the internal and less understood
factors that influence the Mg/Ca ratio. By working with already
published datasets, we were able to simultaneously avoid costs for new
high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
examples, and to also provide pilot-data for areas of existing research
interest. Generally establishing the usefulness and reliability of Mg/Ca
ratios as SST proxy in limpets provides a platform for future research
and an important steppingstone to a better understanding of elemental
ratios in other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with
research background information can be found in Table
\hyperref[Table_1]{1}. Here we will briefly summarise their contextual
information, which can be accessed in more detail at the respective
studies \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While
those studies also included other specimens, their accessibility or
state of preservation did not allow them to be re-analysed.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaurata} & Cambaceres Bay &
ND-1016-3 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & - Tierra del & ND-1016-4 & analysis \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & & analysis; \\
& & & & & Mg, Li, Sr, Ca. \\
& & & & & \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O \\
& & & Westray (UK) & QG1-7189-2 & and \(\delta\)\textsuperscript{13}C \\
& & & & QG2-1061-1 & analysis \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

Specimens were sourced from two main locations: the Beagle Channel in
Tierra del Fuego (Argentina) and the island of West Ray in the Orkney
archipelago of Scotland (UK). Four modern specimens were collected from
the Beagle Channel in Cambaceres Bay in October 2016. The area receives
around 570 mm of precipitation, but the channel waters are predominantly
influenced by marine currents and in the period of October 2015 to
October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units)
\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
February (Figure~\ref{fig-SSTs} -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009 and has been previously analysed for
\(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
values, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray lies about
70 km north of the Scottish mainland and experiences virtually no
freshwater input with salinity values of 34.0 -- 34.5 psu
\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
13.8ºC in August (Figure~\ref{fig-SSTs} -- blue line).

@@

\includegraphics{fig-SSTs-1.pdf}

@@

\caption{\label{fig-SSTs}Monthly Sea Surface Temperature values averaged
from years. SST values were sourced from \citet{Good2020-nl}.}

@@

We also studied 7 archaeological specimens of \emph{P. vulgata} from the
Quoygrew site on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly (810 -- 1229 CE)
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long lived for their size (QG1-7246-1, 12 years within 16
mm of growth) making them particularly susceptible to within-shell
time-averaging because of slow growth rates. That said, the majority are
under 5 years old yielding sufficiently fast growth to nearly capture
seasonal amplitudes of SST.

@@

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz
Zentrum für Archäologie (Mainz-Germany) following previously published
methods \citep{Hausmann2023-ih}. The imaging involves the ablation of
the shell section using an infrared (1064 nm) laser (1--2 mJ, 100 Hz)
onto a surface area of 20--30 µm to generate a plasma plume. This plume
emits light, which is measured using a synchronised spectrometer (delay
time = 0.5 µs, integration time = 1 µs). The resulting light spectrum
quantifies the emission lines of Magnesium (MgII; 279.553 nm) and
Calcium (CaII; 315.887 nm) to determine their intensity ratio. While
these two peaks alone do not represent the molar concentrations of both
elements (as opposed to calibration free LIBS; e.g.
\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa} and can be
used as a reliable indicator of Mg/Ca variation within the shell
carbonate. All values reported here are in arbitrary units based on the
ratio of intensities of the peaks above.

@@

High-quality \(\delta\)\textsuperscript{18}O values were acquired from
previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaurata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the methodology for the stable isotope analyses can be found. In short,
all shells were sectioned along the maximum growth axis to expose the
internal growth structure. From the exposed section carbonate powder
samples were acquired using a micromill with resolutions of 100--200 µm
and by targeting the calcitic M+2 layer of the shells. Analytical
precision of the isotope ratios was consistently at 0.05--0.10‰ and
values are reported in per mil units (‰) relative to the VPDB (Vienna
Pee Dee Belemnite) standard.

@@

\includegraphics{fig-Nac_iso-1.png}

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species (\emph{Nacella} spp., and \emph{Patella
vulgata}, respectively). Due to the high-resolution sampling, all shells
show seasonally resolved and multi-year quasi-sinusoidal records of SST
change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
vulgata} living regularly over 3 years. In those instances where longer
lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
reduced from more than a dozen of samples per year, to only around 7.

@@

\includegraphics{fig-Pat_iso-1.png}

@@

@@

Because it is not possible to directly associate the exact sample
locations used in the analysis of oxygen isotope ratios with the LIBS
data, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy datasets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment (see code in \citep{Hausmann_2024OSF}). The process involved
iterative adjustments to minimise the overall distance between
corresponding points in the two data sets.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating high temperature periods and low ratios
indicating low temperature periods. As examples,
Figure~\ref{fig-Pat_LIBS} shows the 2D-distribution of elemental ratios
across the entire section of ORK-LT5 and the preserved anterior side of
QG1-7188-1 (additional maps can be found in the supplementaries). The
repeating patterns are found across the calcitic layers that are
exterior to the myostracum (layers m+2 and m+3) as well as the layers
that are interior (m-2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
for a detailed explanations of microstructural layers in \emph{P.
vulgaata}). The layer m-1, which is also interior, consists of aragonite
and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
ratio consistently falls below the range of interest (0.2 and above).
Compared to other patelloid shells, some of whose interior is almost
entirely made of aragonite, this layer is very thin. \newline

@@

\includegraphics{fig-Pat_LIBS-1.pdf}

@@

\caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
analysed specimens. (\emph{a+c}) ORK-LT and (\emph{b}+\emph{d})
QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not
identical but are chosen arbitrarily to increase contrast. Areas in grey
consist of low-Mg aragonite and fall outside the chosen colour-range.}

@@

\includegraphics{fig-Pat_Comp-1.png}

@@

\includegraphics{fig-Pat_Corr-1.png}

@@

\caption{\label{fig-Pat_Corr}Correlation graphs for \emph{Patella
vulgata} specimens.}

>>>>

\usepackage{lineno}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Palaeo3}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

@@

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={CONICET-CADIC},addressline={AV. B.
Houssay,
200.},city={Ushuaia},country={Argentina},countrysep={,},postcode={9410},postcodesep={}}

@@

\begin{abstract}
This study provides a reassessment of magnesium to calcium (Mg/Ca)
ratios in Atlantic limpet shells to determine past sea surface
temperatures (SST). While \emph{Patella depressa} along the Spanish
shoreline and \emph{Patella caerulea} in the Mediterranean have
repeatedly produced reliable correlations between SST and Mg/Ca ratios,
this relationship is not the case for other patelloid species.
Particularly, \emph{Patella vulgata} but also \emph{Nacella deaurata}
have been studied using Mg/Ca ratios with mixed or contrary results. In
this study, we present elemental maps of these two species as well as
\emph{Nacella magellanica} together with oxygen isotope ratios
(\(\delta\)\textsuperscript{18}O) that confirm a good relationship with
SST. Our dataset also reassesses a specimen which was previously
unsuccessful in providing significant correlations between
\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. By reassessing
these species and including modern and archaeological specimens (n=12)
from three patelloid species (\emph{P. vulgata}, \emph{N. deaurata}, and
\emph{N. magellanica}) we further add to the growing set of evidence for
the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
a result, these species can in the future serve to determine ontogenetic
age and season of capture as well as to reveal locations of interest
within the growth record (i.e.~annual temperature minima and maxima) for
targeted \(\delta\)\textsuperscript{18}O and clumped isotope analysis.
\end{abstract}

@@

Limpet shells are commonly found within archaeological sites and past
shorelines due to their robust carbonate structure and the long-term use
of limpets as a marine food source
\citep[\citet{Robson2023-oz}]{Alvarez2011-rm}. They have been
successfully studied in the context of coastal subsistence economies and
site occupation on seasonal
\citep{Shackleton1973-ij, Parker2020-oy, Bosch2018-ud} or long-term
scales \citep{Ortiz2015-mr}, as well as archives of palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change recorded in
patelloid limpet shells largely relies on the measurement of
\(\delta\)\textsuperscript{18}O values within the calcitic parts of the
shell, but attempts have been made to also use elemental ratios, such as
magnesium to calcium (Mg/Ca) ratios
\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative measure
that potentially provides a more accurate SST estimate than
\(\delta\)\textsuperscript{18}O-values,as the latter proxy is affected
by changes in the mixing of fresh- and saltwater. In addition, the data
acquisition of elemental ratios --- either through
laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
induced breakdown spectroscopy (LIBS), can be much faster and
cost-effective, increasing the number of specimens that can be studied
overall \citep{Durham2017-fh, Hausmann2023-ih}.

@@

Recent research particularly of \emph{Patella} spp. in the Mediterranean
and southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
However, there remains a lack of clarity for Atlantic limpet species,
particularly since it was shown, that the Mg/Ca ratio in their shells
are not reliable recorders of palaeotemperature \citep{Graniero2017-io}.
Using modern reference material Graniero \emph{et al}.
\citeyearpar{Graniero2017-io} tested multiple elemental ratios (Mg/Ca,
Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with
no ratios showing correlations with corresponding SST changes nor any
sinusoidal patterns that would indicate a seasonal environmental control
of elemental ratios. This study included two species also used in our
study (\emph{Patella vulgata}, \emph{Nacella deaurata}), with one
\emph{P. vulgata} specimen (i.e.~specimen ORK-LT5) being analysed by
both. We also added specimens of the species \emph{Nacella magellanica,}
which --- similar to the other two species --- is also an established
palaeothermometer using \(\delta\)\textsuperscript{18}O-values
\citep{Colonese2012-ct, gutiérrez-zugasti2017, Nicastro2020-ih}.

In this study, we repeat and expand upon the analysis of Atlantic
limpets to determine the reliability of Mg/Ca ratios as
palaeotemperature proxies. To do this, we sampled a set of previously
published and unpublished limpet specimens dating to modern and
archaeological contexts using LIBS. LIBS allows us to carry out 2D
imaging of entire shell sections, which facilitates navigation of the
complex elemental structure of the shell and a better separation of the
external, temperature-related changes from the internal and less
understood factors that influence the Mg/Ca ratio. By working with
already published datasets, we were able to simultaneously avoid costs
for new high-resolution \(\delta\)\textsuperscript{18}O-data, to use
real-world examples, and to also provide pilot-data for areas of
existing research interest. Generally establishing the usefulness and
reliability of Mg/Ca ratios as SST proxy in limpets provides a platform
for future research and an important steppingstone to a better
understanding of elemental ratios in other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with
research background information can be found in Table
\hyperref[Table_1]{1}. Here we will briefly summarise their contextual
information, which can be accessed in more detail from the respective
studies \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While
those studies also included other specimens, their accessibility or
state of preservation did not allow them to be re-analysed.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1389}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2361}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaurata} & Cambaceres Bay &
ND-1016-3 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & - Tierra del & ND-1016-4 & analysis \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & & analysis; \\
& & & & & Mg, Li, Sr, Ca \\
& & & & & \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & QG1-7189-2 & analysis \\
& & & & QG2-1061-1 & \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

Specimens were sourced from two main locations: the Beagle Channel in
Tierra del Fuego (Argentina) and the island of Westray in the Orkney
archipelago of Scotland (UK). Four modern specimens were collected from
the Beagle Channel in Cambaceres Bay in October 2016. The area receives
around 570 mm of annual precipitation, but the channel waters are
predominantly influenced by marine currents and in the period of October
2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity
units) \citep{Nicastro2020-ih}. SSTs range from 30.5ºC in August to
31.9ºC in February (Figure~\ref{fig-SSTs} -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009 and has been previously analysed for
\(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
values, as well as for elemental ratios via LA-ICP-MS
\citep{Graniero2017-io}. Westray lies about 70 km north of the Scottish
mainland and experiences virtually no freshwater input with salinity
values of 34.0 -- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges
from 34.4ºC in March to 35.3ºC in August (Figure~\ref{fig-SSTs} -- blue
line).

@@

\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}

@@

\caption{\label{fig-SSTs}Monthly sea surface temperature values averaged
from 2007 to 2023. Values were sourced from \citet{Good2020-nl}.}

@@

We also studied 7 archaeological specimens of \emph{P. vulgata} from the
Quoygrew site on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly (810 -- 1229 CE)
\citep{Surge2012-ba}.

@@

@@

High-quality \(\delta\)\textsuperscript{18}O values were acquired from
previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaurata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the methodology for the stable isotope analyses can be found. In short,
all shells were sectioned along the maximum growth axis to expose the
internal growth structure. From the exposed section carbonate powder
samples were acquired using a micromill with resolutions of 100--200 µm
and by targeting the calcitic layers of the shells (m+2 and m+3, see
\citep{Fenger2007-gf}). The exact location of the sampled carbonate
powder is not available apart from the distance to the growth edge,
where reported. Analytical precision of the isotope ratios was
consistently at 0.05--0.10‰ and values are reported in per mil units (‰)
relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species (\emph{Nacella} spp., and \emph{Patella
vulgata}, respectively). Due to the high-resolution sampling, all shells
show seasonally resolved and multi-year quasi-sinusoidal records of SST
change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
vulgata} living regularly over 3 years. In those instances where longer
lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
reduced from more than a dozen of samples per year, to only around 7.
Together they show a range of lifespans, with some comparatively long
lived for their size (QG1-7246-1, 12 years within 16 mm of growth)
making them particularly susceptible to within-shell time-averaging
because of slow growth rates. That said, the majority yielded
sufficiently fast growth to nearly capture seasonal amplitudes of SST
with ages of 2--3 years (n=3), 3--4 years (n=2), and 5 years (n=2).

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.png}

@@

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz
Zentrum für Archäologie (Mainz-Germany) following previously published
methods \citep{Hausmann2023-ih}. The imaging involves the ablation of
the shell section using an infrared (1064 nm) laser (1---2 mJ, 100 Hz)
onto a surface area of 20---30 µm to generate a plasma plume. This plume
emits light, which is measured using a synchronised spectrometer (delay
time = 0.5 µs, integration time = 1 µs). The resulting light spectrum
quantifies the emission lines of magnesium (MgII: 279.553 nm) and
calcium (CaII: 315.887 nm) to determine their intensity ratio. While
these two peaks alone do not represent the molar concentrations of both
elements (as opposed to calibration free LIBS; e.g.
\citep{Martinez-Minchero2023-pq}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa} and can be
used as a reliable indicator of Mg/Ca variation within the shell
carbonate. All values reported here are in arbitrary units based on the
ratio of intensities of the peaks above.

@@

Because it is not possible to directly associate the exact sample
locations used in the analysis of oxygen isotope ratios with the LIBS
data, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy datasets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This can involve selecting
appropriate distance measures and constraints to ensure meaningful
alignment. We used the default settings for symmetric step patterns and
euclidean distance measures (see code in Hausmann et
al.\citeyearpar{hausmann2024confirmation}).

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating high temperature periods and low ratios
indicating low temperature periods. As examples,
Figure~\ref{fig-Pat_LIBS} shows the 2D-distribution of elemental ratios
across the entire section of ORK-LT5 and the preserved anterior side of
QG1-7188-1 (additional maps can be found in the supplementaries). The
repeating patterns are found across the calcitic layers that are
exterior to the myostracum (layers m+2 and m+3) as well as the layers
that are interior (m--2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
for a detailed explanations of microstructural layers in \emph{P.
vulgata}). The layer m-1, which is also interior, consists of aragonite
and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
ratio consistently falls below the range of interest (0.2 and above).
Compared to other patelloid shells, some of whose interior is almost
entirely made of aragonite, this layer is very thin. \newline

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}

@@

\caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
analysed specimens. (\emph{a+c}) ORK-LT and (\emph{b}+\emph{d})
QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not
identical but are chosen arbitrarily to increase contrast. Areas in grey
consist of low-Mg aragonite and fall outside the chosen colour-range.
White dashed lines indicate location of line scans.}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.png}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.png}

@@

\caption{\label{fig-Pat_Corr}Correlation graphs for Mg/Ca and
\(\delta\)\textsuperscript{18}O values in \emph{P. vulgata} specimens.}
